A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-01 paper.tex

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence

Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
% TODO rephrase this, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

\begin{figure}
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (3, 1) {};
\node (s4) [greynode] at (1, 2) {};
\node (s5) [greynode] at (5, 3) {};
\node (s6) [greynode] at (3, 4) {};

% \node [anchor=north west] at (0,6) {A};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

% \node [anchor=north west] at (9,6) {B};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

% \node [anchor=north west] at (9,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG and a concrete
encoding of the information. Node \noderef{d} is a recombination
event as it is associated with a non-null breakpoint, and has
two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
common ancestor events.
}
\end{figure}

In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

>>>>

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we record the intervals of inheritance
between genomes (and not causal \emph{events}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node \noderef{e}
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that \noderef{e} the ancestor of \noderef{a} over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
% TODO rephrase this, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

In keeping with its original derivation in terms of a stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an
ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
% is usually concretely encoded by associating a recombination breakpoint
% (which is null for common ancestor events) and an ordered list of parents
% with each node. We refer to such a  data structure as an
% ``event ARG'' (eARG).
% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.
